On joint estimation of ancestral states and phylogenetic model parameters

### 2017-10-30 joint_inf.tex
forgot a scaling factor!
add new optimization plot

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{topology-inconsistency-legend}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
%EM Not sure what's being suppressed here. Sounds naughty.
%das: I meant suppressing terms not involving $w'$, but I guess I already did that...
Simplifying, the likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
%EM Rather confusing to be using ' for two things here. What about using a dot over the l? In retrospect something other than ' would have been good because of the proliferation of w^{'2} etc. Prob not worth changing now though.
The derivative of the log likelihood is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
%EM I don't think you mean "any inequality"... what about sin(ell') > 0?
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
%EM Is this obvious at this point? Couldn't b^2 - 4ac be zero in principle?
%das: only if x or y is zero (though one could be nonzero).
%b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero.
%c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
%Added clarification.
Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{branch-length-inconsistency-legend}
\caption{Region of inconsistency with $x$ and $y$ fixed}
\label{fig:bl-inconsistency}
\end{figure}

@@

@@

When computing the global maximum for $0 \le x',y',w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}'$ in Fig.~\ref{fig:w_hat_joint}.
The region in which $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is now much smaller, but still nontrivial.
Additionally, the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}'$ scattered throughout the $(x,y)$ lattice.
Even given the correct Farris topology we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{w_hat_heatmap_joint}
\caption{
    Estimating $(\hat{x}', \hat{y}', \hat{w}')$.
}
\label{fig:w_hat_joint}
\end{figure}

>>>>

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=.9\textwidth]{topology-inconsistency-legend}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
%EM Not sure what's being suppressed here. Sounds naughty.
%das: I meant suppressing terms not involving $w'$, but I guess I already did that...
Simplifying, the likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = \frac{1}{8} (1+x^2+y^2+4xy^2+x^2y^2),
$$
$$
\alpha_2 := \alpha_2(x, y) = \frac{1}{8}(1+x^2+y^2-4xy^2+x^2y^2),
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
%EM Rather confusing to be using ' for two things here. What about using a dot over the l? In retrospect something other than ' would have been good because of the proliferation of w^{'2} etc. Prob not worth changing now though.
The derivative of the log likelihood is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
%EM I don't think you mean "any inequality"... what about sin(ell') > 0?
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w') \lessgtr 0$, we equivalently consider $f(w') \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
%EM Is this obvious at this point? Couldn't b^2 - 4ac be zero in principle?
%das: only if x or y is zero (though one could be nonzero).
%b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero.
%c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
%Added clarification.
Excepting all the cases where $x=0$ or where $y=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
\includegraphics[width=.9\textwidth]{branch-length-inconsistency-legend}
\caption{Region of inconsistency with $x$ and $y$ fixed}
\label{fig:bl-inconsistency}
\end{figure}

@@

\subsubsection{Restricted general limiting case}

Let $x=y=u$ and $x'=y'=w'=u'$, a ``restricted'' Farris topology.
We will show that, as $u\rightarrow 0$, $\hat{u'}\rightarrow u_0 > 0$, which is less than ideal.
Our first mode of approximation is to let $u'^p\approx 0$ for all $p > 1$, preserving only linear terms.
After some algebra, we show the likelihood is equal to
\begin{align*}
    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
    &=        p_{\emptyset}  \cdot\log(1+2u'^2+4u'^3+u'^4) \\
    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log(1-u'^4) \\
    &\qquad + (p_{12}+p_{23})         \cdot\log(1-2u'^2+u'^4) \\
    &\qquad + p_{13}         \cdot\log(1+2u'^2-4u'^3+u'^4) \\
    &\qquad - \log(8) \\
    &\qquad + p_{\emptyset}  \cdot\log((1+u')^5)\\
    &\qquad + p_{1}          \cdot\log((1+u')^4(1-u'))\\
    &\qquad + p_{2}          \cdot\log((1+u')^4(1-u'))\\
    &\qquad + p_{3}          \cdot\log((1+u')^4(1-u'))\\
    &\qquad + p_{12}         \cdot\log((1+u')^3(1-u')^2)\\
    &\qquad + p_{13}         \cdot\log((1+u')^4(1-u'))\\
    &\qquad + p_{23}         \cdot\log((1+u')^3(1-u')^2)\\
    &\qquad + p_{123}        \cdot\log((1+u')^4(1-u')).
\end{align*}
Expanding and simplifying,
\begin{align*}
    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
    &=        p_{\emptyset}  \cdot\log((1+u')(1-u'+3u'^2+u'^3))\\
    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log(1+u'^2) \\
    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log((1+u')(1-u')) \\
    &\qquad + 2(p_{12}+p_{23})         \cdot\log((1+u')(1-u')) \\
    &\qquad   p_{13}  \cdot\log((1-u')(1+u'+3u'^2-u'^3))\\
    &\qquad - \log(8) \\
    &\qquad + 5\cdot\log(1+u')\\
    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u').
\end{align*}
Now, applying $u'^p\rightarrow 0$ for $p > 1$, this approaches
\begin{align*}
    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
    &\rightarrow        p_{\emptyset}  \cdot\log((1+u')(1-u'))\\
    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123})          \cdot\log((1+u')(1-u')) \\
    &\qquad + 2(p_{12}+p_{23})         \cdot\log((1+u')(1-u')) \\
    &\qquad   p_{13}  \cdot\log((1-u')(1+u'))\\
    &\qquad - \log(8) \\
    &\qquad + 5\cdot\log(1+u')\\
    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u'),
\end{align*}
and
\begin{align*}
    \ell_{\tau_1,\{u\}}(\tau_1, \{u'\}; \fullAncestralSplitPartitions(\tau_1,\{u'\}))
    &\rightarrow (1+p_{12}+p_{23})\cdot\log((1+u')(1-u'))\\
    &\qquad - \log(8) \\
    &\qquad + 5\cdot\log(1+u')\\
    &\qquad - (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1+u') \\
    &\qquad + (p_{1}+p_{2}+p_{3}+p_{123}+2p_{12}+2p_{23})\cdot\log(1-u').
\end{align*}
This is of the form
$$
\alpha\log(1+u') + \beta\log(1-u')
$$
and maximized at $\hat{u}' = (\alpha - \beta) / (\alpha + \beta)$ provided $\hat{u}' \in [0,1]$.
As $u\rightarrow 0$, $\alpha\rightarrow 21/4$ and $\beta\rightarrow 9/4$ so $\hat{u}'\rightarrow 2/5.$
I don't know if this is interesting yet.

\subsubsection{General general case}

@@

When seeking a global maximum for $0 \le x',y',w' \le 1$, we use the method of differential evolution \cite{Storn1997}---we obtain the fit $\hat{w}'$ in Fig.~\ref{fig:w_hat_joint}.
The estimates for $\hat{w}'$ when $x$ and $y$ are large seem to be unbiased, with $\hat{w}'$ increasing as $y$ increases, though the region where $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is similar to that in the restricted case.
Although the restricted and general case are similar, comparing Figures~\ref{fig:bl-inconsistency} and~\ref{fig:w_hat_joint} shows that the general case has a smaller region where $\hat{w}'$ is one, as expected.
Nevertheless, this region encompasses the majority of the parameter space; even given the correct Farris topology and performing our best possible optimization, we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --general-branch-lengths --delta .001 --plot-name figures/w-hat-heatmap-joint-diff-evol.svg --n-jobs 10
\includegraphics[width=.9\textwidth]{w-hat-heatmap-joint-diff-evol}
\caption{
    Estimating $(\hat{x}', \hat{y}', \hat{w}')$.
}
\label{fig:w_hat_joint}
\end{figure}
